Practical Machine Learning Course Project
===========================

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).
2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details. 

```{r setup, message=FALSE, warning=FALSE}
library("knitr")
library("ggplot2")
library("reshape")
library("gridExtra")
library("RColorBrewer")
library("caret")

opts_chunk$set(cache=TRUE, message=FALSE, warning=FALSE, fig.height=5, fig.width=12)
```

Load in the data.  

```{r loaddata, cahce=TRUE}
setwd("/Users/kwalker/git_projects/PredMod_CourseProject/")
raw_test <- read.csv("pml-testing.csv", na.strings=c("NA", "#DIV/0", ""), stringsAsFactors=FALSE)
raw_train <- read.csv("pml-training.csv", na.strings=c("NA", "#DIV/0", ""), stringsAsFactors=FALSE)
```

Clean the data:  
1. Remove variables that are mostly NAs   
2. Remove first 6 descriptors (user_name, etc.)  
3. Change format to numeric for all but classe   
4. Remove collinear variables    

```{r cleandata}
clean <- raw_train

# transform: remove variables that are mostly NA
clean.cut <- clean[,colSums(is.na(clean)) == 0] 

# transform: remove first 6 descriptors
clean.cut <- clean.cut[ ,-c(1:6)]

# make numeric
for (i in 1:53 ) clean.cut[,i] <- as.numeric(clean.cut[,i])

# remove collinear variables 
co <- cor(clean.cut[ ,1:53], use="pairwise.complete.obs")
diag(co) <- 0
pairs <- NULL
for ( i in 1:length(co[,1]) ) {
     for ( x in 1:length(co[1,])) {
          if (!is.na(co[x,i]) & co[x,i] > .9) {
               z <- i + 1
               m <- x + 1
               rname <- row.names(co[i:z,])[1]
               cname <- colnames(co[,x:m])[1]
               pairs <- rbind(pairs, c(rname, cname, co[x,i]))
          }
     }
}
pairs <- data.frame(pairs, stringsAsFactors=FALSE)
pairs <- pairs[order(-as.numeric(pairs$X3)),]
print(pairs)
collinears <- c("total_accel_belt", "gyros_dumbell_z", "accel_belt_y")
clean.cut <- clean.cut[ , !(colnames(clean.cut) %in% collinears)]
```

Perform the same transformations on test.  

```{r cleanTest}
cleanTest <- raw_test
# remove first six descriptors
cleanTest <- cleanTest[ ,-c(1:6)]

# make numeric
for (i in c(1:153) ) cleanTest[,i] <- as.numeric(cleanTest[,i])

# remove variables that are mostly NAs
cleanTest <- cleanTest[ ,colSums(is.na(cleanTest)) == 0]

# remove collinear variables
cleanTest <- cleanTest[ , !(colnames(cleanTest) %in% collinears)]
```

Split the train data into two parts.  

```{r splitdata}
inTrain <- createDataPartition(y=clean.cut$classe, p=.75, list=FALSE)
training <- clean.cut[inTrain, ]
testing <- clean.cut[-inTrain, ]
```

